---
title: "Take-home Exercise 2: VAST Challenge 2025 - Mini Challenge 3"
author: "Vanessa Riadi"
date: "May 26, 2025"
date-modified: "last-modified"
format:
  html:
    toc: true
    prefer-html: true
execute: 
  echo: true
  eval: true
  warning: false
  freeze: true
---

# [1]{style="color:#d496d4; background:none;"} Overview

Clepper Jessen, an investigative journalist, is uncovering possible corruption in Oceanus linked to the closure of Nemo Reef and the arrival of pop star Sailor Shift. Using intercepted radio communications, a knowledge graph was created to map out interactions between people, vessels, and organizations. This Mini Challenge aims to support Clepper's investigation through data-driven visual analytics, uncovering patterns, pseudonym usage, social groupings, and suspicious behavior—especially concerning Nadia Conti, a previously implicated figure in illegal fishing.

# [2]{style="color:#d496d4; background:none;"} Objective

The Mini Challenge 3 tasked us to design and develop up to three graph-based data visualizations that:

-   Identify temporal patterns in communications.

-   Reveal relationships and influence among entities (people, vessels, groups).

-   Expose pseudonym usage and investigate Nadia Conti's activity for potential misconduct.

This is part 1 of the Mini Challenge 3 where we will be looking at question 1:

a)  Develop a graph-based visual analytics approach to identify any daily temporal patterns in communications.

b)  How do these patterns shift over the two weeks of observations?

c)  Focus on a specific entity and use this information to determine who has influence over them.

and 3c) How does your understanding of activities change given your understanding of pseudonyms?

::: {.nursebox .nurse data-latex="nurse"}
For part 2 go to Hoa's page (link pending)

For part 3 go to Summer's page (link pending)
:::

# [3]{style="color:#d496d4; background:none;"} Analytical Toolkit: RStudio

RStudio and Quarto are used as the primary analytical tools for this challenge. The data will be analyzed and visualized using the **tidyverse** suite and advanced network visualization packages to explore the knowledge graph and temporal communication dynamics.

Before we get started, it is important for us to ensure that the required R packages have been installed.

::: callout-note
# Install [**pacman**](https://cran.r-project.org/web/packages/pacman/index.html){target="_blank"} package

If you have yet to install pacman, install itby typing below in the Console:

```{r}
#| eval: false
options(repos = c(CRAN = "https://cloud.r-project.org"))
install.packages("pacman")
```
:::

We then load the following R packages using the `pacman::p_load()` function:

```{r}
pacman::p_load(tidyverse, tidygraph, ggraph, jsonlite, ggplot2,
               SmartEDA, lubridate, ggthemes, readr, readxl, knitr, dplyr, visNetwork)
```

# [4]{style="color:#d496d4; background:none;"} Data

The core dataset is a knowledge graph derived from radio communications intercepted over a two-week period on Oceanus. Each node in the graph represents an entity such as a person, vessel, pseudonym, or organization. Edges represent interactions or co-occurrences in communications. Edge attributes include timestamp, topic, message type, and possible pseudonym use.

Supplementary information from the story given on the Mini Challenge page includes:

-   Entity roles and affiliations (e.g., Green Guardians, Sailor Shift's crew).

-   Known pseudonyms (e.g., “Boss”, “The Lookout”).

-   Historical involvement in illicit activities (e.g., Nadia Conti).

Refer to the data description details in [Appendix A](#appendixa)

## [4.1]{style="color:#d496d4; background:none;"} Load the Data

First we load the data.

```{r}
graph_data <- fromJSON("data/MC3_graph.json")
schema_data <- fromJSON("data/MC3_schema.json")
```

## [4.2]{style="color:#d496d4; background:none;"} Check the Data

```{r}
glimpse(graph_data)
```

```{r}
glimpse(schema_data)
```

::: {.nursebox .nurse data-latex="nurse"}
#### **Observation** {.no-top-padding}

-   There are a lot of NA values because only certain types have certain information. Let's look more towards each edges and nodes.
-   Since the focus is recorded intercepted radio communications over the last two weeks, we will also focus on Communications records.
-   We also need to clean and convert some of the data types (e.g Convert timestamp from chr to datetime format)

But let's explore more first!
:::

## [4.3]{style="color:#d496d4; background:none;"} Extracting the edges and nodes tables

Next, `as_tibble()` of **tibble** package package is used to extract the nodes and links tibble data frames from *mc3* tibble dataframe into two separate tibble dataframes called *mc3_nodes* and *mc3_edges* respectively.

```{r}
nodes_tbl <- as_tibble(graph_data$nodes)
edges_tbl <- as_tibble(graph_data$edges)
```

```{r}
head(nodes_tbl)
```

```{r}
head(edges_tbl)
```

### View distinct node types

```{r}
table(nodes_tbl$type)
```

### [4.3.1]{style="color:#d496d4; background:none;"} Initial EDA

In the code chunk below, `ExpCatViz()` of SmartEDA package is used to reveal the frequency distribution of all categorical fields in *nodes_tbl* dataframe.

```{r}
ExpCatViz(data=nodes_tbl,
          col="lightblue")
```

::: callout-note
What useful discovery can you obtained from the visualisation above?
:::

On the other hands, code chunk below uses `ExpCATViz()` of SmartEDA package to reveal the frequency distribution of all categorical fields in *edges_tbl* dataframe.

```{r}
ExpCatViz(data=edges_tbl,
          col="lightblue")
```

::: callout-note
What useful discovery can you obtained from the visualisation above?
:::

## [4.4]{style="color:#d496d4; background:none;"} Data Preparation {#dataprep}

We will performs the following data cleaning and data wrangling tasks:

-   Extract communication events with timestamps
-   Clean timestamp
-   Extract sender and receiver links
-   Separate senders and receivers
-   Join sender-receiver pairs
-   Join with communication details
-   Join to get sender/receiver labels
-   Save the result for use

### [4.4.1]{style="color:#d496d4; background:none;"} Extract communication events with timestamps

```{r}
comm_events <- nodes_tbl %>%
  filter(type == "Event", sub_type == "Communication")

head(comm_events)
```

### [4.4.2]{style="color:#d496d4; background:none;"} Extract necessary fields

We'll use `id`, `timestamp` and `content`

```{r}
comm_events <- comm_events %>%
  select(id, timestamp = timestamp, content = content)

head(comm_events)
```

### [4.4.3]{style="color:#d496d4; background:none;"} Convert timestamp from chr to datetime format

```{r}
comm_events <- comm_events %>%
  mutate(timestamp = ymd_hms(timestamp))

head(comm_events)
```

### [4.4.3]{style="color:#d496d4; background:none;"} Filter edges for sender and receiver

```{r}
comm_edges <- edges_tbl %>%
  filter(type %in% c("sent", "received")) %>%
  select(source, target, type)

head(comm_edges)
```

### [4.4.4]{style="color:#d496d4; background:none;"} Separate and rename

```{r}
senders <- comm_edges %>%
  filter(type == "sent") %>%
  rename(sender = source, message_id = target)

receivers <- comm_edges %>%
  filter(type == "received") %>%
  rename(receiver = target, message_id = source)
```

### [4.4.5]{style="color:#d496d4; background:none;"} Join sender and receiver info

```{r}
comm_pairs <- inner_join(senders, receivers, by = "message_id")

head(comm_pairs)
```

### [4.4.6]{style="color:#d496d4; background:none;"} Combine with communication details

```{r}
comm_full <- comm_pairs %>%
  left_join(comm_events, by = c("message_id" = "id"))

head(comm_full)
```

### [4.4.7]{style="color:#d496d4; background:none;"} Join entity labels for sender and receiver

```{r}
entity_labels <- nodes_tbl %>%
  filter(type == "Entity") %>%
  select(id, label, sub_type)

head(entity_labels)
```

### [4.4.7]{style="color:#d496d4; background:none;"} Combine everything and save to CSV

```{r}
comm_full <- comm_full %>%
  left_join(entity_labels, by = c("sender" = "id")) %>%
  rename(sender_label = label, sender_type = sub_type) %>%
  left_join(entity_labels, by = c("receiver" = "id")) %>%
  rename(receiver_label = label, receiver_type = sub_type)

head(comm_full)
```

For the purpose of this Take Home Exercise, we will extract out to CSV so the rest of the group can work with this cleaned data!

```{r}
#| eval: false
write_csv(comm_full, "data/communications_full.csv")
```

# [5]{style="color:#d496d4; background:none;"} Data Visualisation, Observation, and Insights

For part 1 of the Mini Challenge 3 Let's look specifically into event nodes where communications occur.

## [5.1]{style="color:#d496d4; background:none;"} Develop a graph-based visual analytics approach to identify any daily temporal patterns in communications.

How do these patterns shift over the two weeks of observations?

```{r}
comm_full %>%
  mutate(date = as.Date(timestamp)) %>%
  count(date) %>%
  ggplot(aes(x = date, y = n)) +
  geom_line(color = "steelblue", size = 1.2) +
  labs(
    title = "Daily Communication Volume",
    x = "Date", y = "Number of Messages"
  ) +
  theme_minimal()
```

::: {.nursebox .nurse data-latex="nurse"}
#### 🎀 **Insights Plot 1** {.no-top-padding}

-   
:::

## [5.2]{style="color:#d496d4; background:none;"} Focus on a specific entity and use this information to determine who has influence over them.

```{r}
# Create edge list
graph_edges <- comm_full %>%
  count(sender_label, receiver_label) %>%
  filter(!is.na(sender_label), !is.na(receiver_label)) %>%
  rename(from = sender_label, to = receiver_label, value = n)

# Create node list from edges
graph_nodes <- tibble(name = unique(c(graph_edges$from, graph_edges$to))) %>%
  mutate(id = name, label = name, group = name)

# visNetwork expects numeric node IDs, but we can use labels for both
visNetwork(nodes = graph_nodes, edges = graph_edges) %>%
  visEdges(arrows = "to") %>%
  visOptions(highlightNearest = TRUE, nodesIdSelection = TRUE) %>%
  visLayout(randomSeed = 123) %>%
  visPhysics(stabilization = TRUE) %>%
  visInteraction(navigationButtons = TRUE) %>%
  visLegend()

```
::: {.nursebox .nurse data-latex="nurse"}
#### 🎀 **Insights Plot 2** {.no-top-padding}

-   
:::

## [5.3]{style="color:#d496d4; background:none;"} Pseudonym Usage & Nadia Conti's Activity

### [5.3.1]{style="color:#d496d4; background:none;"} Filter: Communications involving Nadia Conti

```{r}
# Filter communications with Nadia
nadia_msgs <- comm_full %>%
  filter(sender_label == "Nadia Conti" | receiver_label == "Nadia Conti")

# See who she communicates with
nadia_msgs %>%
  count(sender_label, receiver_label, sort = TRUE)

```

### [5.3.2]{style="color:#d496d4; background:none;"} Create tidygraph object for Nadia's ego network

```{r}
nadia_edges <- nadia_msgs %>%
  count(sender_label, receiver_label) %>%
  rename(from = sender_label, to = receiver_label, weight = n)

nadia_nodes <- tibble(name = unique(c(nadia_edges$from, nadia_edges$to)))

nadia_graph <- tbl_graph(nodes = nadia_nodes, edges = nadia_edges, directed = TRUE)

ggraph(nadia_graph, layout = "kk") +
  geom_edge_link(aes(width = weight), alpha = 0.6, color = "purple") +
  geom_node_point(color = "orange", size = 6) +
  geom_node_text(aes(label = name), repel = TRUE, size = 4) +
  labs(title = "Nadia Conti's Communication Network") +
  theme_void()
```

::: {.nursebox .nurse data-latex="nurse"}
#### 🎀 **Insights Plot 3** {.no-top-padding}

-   
:::

# [6]{style="color:#d496d4; background:none;"} Reference

\- [**VAST 2024 Mini Challenge 3**](https://vast-challenge.github.io/2025/MC3.html){target="_blank"}\
- [**jsonlite**](https://cran.r-project.org/package=jsonlite){target="_blank"}: A robust and high-performance JSON parser and generator for R, useful for working with web APIs and structured data.  \
- [**tibble**](https://cran.r-project.org/package=tibble){target="_blank"}: A modern reimagining of data frames in R that prints cleaner and is better for programming.\
- [**dplyr**](https://cran.r-project.org/package=dplyr){target="_blank"}: A grammar of data manipulation, making data wrangling concise, fast, and readable.\
- [**knitr**](https://cran.r-project.org/package=knitr){target="_blank"}: A dynamic report generation tool in R, used for rendering markdown reports and tables.\
- [**SmartEDA**](https://cran.r-project.org/package=SmartEDA){target="_blank"}: Automatically generates EDA reports including statistical summaries and visualizations.\
- [**lubridate**](https://cran.r-project.org/package=lubridate){target="_blank"}: Simplifies the parsing, manipulation, and extraction of date-time data.\
- [**readr**](https://cran.r-project.org/package=readr){target="_blank"}: Provides fast and friendly functions for reading and writing rectangular data (e.g., CSV).\
- [**ggplot2**](https://cran.r-project.org/package=ggplot2){target="_blank"}: A versatile and widely used R package for creating elegant data visualizations based on the Grammar of Graphics.\
- [**visNetwork**](https://datastorm-open.github.io/visNetwork/){target="_blank"}: Enables interactive network visualizations in R using the vis.js JavaScript library.\
- [**tidygraph**](https://cran.r-project.org/package=tidygraph){target="_blank"}: Provides a tidy framework for network analysis, integrating graph theory with tidy data principles.\
- [**ggraph**](https://cran.r-project.org/package=ggraph){target="_blank"}: Extends ggplot2 to visualize network and graph structures with customizable layouts.\
- [**ggrepel**](https://cran.r-project.org/package=ggrepel){target="_blank"}: Automatically adjusts overlapping text labels in ggplot2 and ggraph visualizations to improve clarity.

# [7]{style="color:#d496d4; background:none;"} Appendix

## [7.1]{style="color:#d496d4; background:none;"} Appendix A {#appendixa}

![](data/MC3_data_description.pdf){fig-align="center" width="700" height="350"}
