---
title: "VAST Challenge 2025 - Mini Challenge 3 Question 1"
author: "Vanessa Riadi"
date: "May 26, 2025"
date-modified: "last-modified"
format:
  html:
    toc: true
    prefer-html: true
execute: 
  echo: true
  eval: true
  warning: false
  freeze: true
---

# [1]{style="color:#d496d4; background:none;"} Overview

Clepper Jessen, an investigative journalist, is uncovering possible corruption in Oceanus linked to the closure of Nemo Reef and the arrival of pop star Sailor Shift. Using intercepted radio communications, a knowledge graph was created to map out interactions between people, vessels, and organizations. This Mini Challenge aims to support Clepper's investigation through data-driven visual analytics, uncovering patterns, pseudonym usage, social groupings, and suspicious behavior—especially concerning Nadia Conti, a previously implicated figure in illegal fishing.

# [2]{style="color:#d496d4; background:none;"} Objective

The Mini Challenge 3 tasked us to design and develop up to three graph-based data visualizations that:

-   Identify temporal patterns in communications.

-   Reveal relationships and influence among entities (people, vessels, groups).

-   Expose pseudonym usage and investigate Nadia Conti's activity for potential misconduct.

This is part 1 of the Mini Challenge 3 where we will be looking at question 1:

a)  Develop a graph-based visual analytics approach to identify any daily temporal patterns in communications.

b)  How do these patterns shift over the two weeks of observations?

c)  Focus on a specific entity and use this information to determine who has influence over them.

and 3c) How does your understanding of activities change given your understanding of pseudonyms?

::: {.nursebox .nurse data-latex="nurse"}
For part 2 go to [Hoa's Netlify page](https://isss608-ay2024-25t.netlify.app/)

For part 3 go to [Summer's Netlify page](https://summernguyenn.netlify.app/)
:::

# [3]{style="color:#d496d4; background:none;"} Analytical Toolkit: RStudio

RStudio and Quarto are used as the primary analytical tools for this challenge. The data will be analyzed and visualized using the **tidyverse** suite and advanced network visualization packages to explore the knowledge graph and temporal communication dynamics.

::: callout-note
# Install [**pacman**](https://cran.r-project.org/web/packages/pacman/index.html){target="_blank"} package

Before we get started, it is important for us to ensure that the required R packages have been installed. If you have yet to install pacman, install it by typing below in the Console:

```{r}
#| code-fold: true
#| eval: false
options(repos = c(CRAN = "https://cloud.r-project.org"))
install.packages("pacman")
```
:::

We then load the following R packages using the `pacman::p_load()` function:

```{r}
#| code-fold: true
pacman::p_load(tidyverse, tidygraph, ggraph, jsonlite, ggplot2,
               SmartEDA, lubridate, ggthemes, readr, readxl, knitr, dplyr, visNetwork)
```

# [4]{style="color:#d496d4; background:none;"} Data

The core dataset is a knowledge graph derived from radio communications intercepted over a two-week period on Oceanus. Each node in the graph represents an entity such as a person, vessel, pseudonym, or organization. Edges represent interactions or co-occurrences in communications. Edge attributes include timestamp, topic, message type, and possible pseudonym use.

Supplementary information from the story given on the Mini Challenge page includes:

-   Entity roles and affiliations (e.g., Green Guardians, Sailor Shift's crew).

-   Known pseudonyms (e.g., “Boss”, “The Lookout”).

-   Historical involvement in illicit activities (e.g., Nadia Conti).

Refer to the data description details in [Appendix A](#appendixa)

## [4.1]{style="color:#d496d4; background:none;"} Load the Data

```{r}
graph_data <- fromJSON("data/MC3_graph.json")
schema_data <- fromJSON("data/MC3_schema.json")
```

## [4.2]{style="color:#d496d4; background:none;"} Check the Data

### [4.2.1]{style="color:#d496d4; background:none;"} graph_data

```{r}
#| code-fold: true
glimpse(graph_data)
```

### [4.2.2]{style="color:#d496d4; background:none;"} schema_data

```{r}
#| code-fold: true
glimpse(schema_data)
```

::: {.nursebox .nurse data-latex="nurse"}
#### **Observation** {#glimpse .no-top-padding}

-   Each data set are already split to nodes and edges thus we will extract the edges and nodes to their own tables.
-   There are a lot of NA values because only certain types have certain information. For example type = "events" and sub_type = "Communications" will have timestamp and content.
-   Since the focus is recorded intercepted radio communications over the last two weeks, we will also focus on Communications records.
-   We also need to clean and convert some of the data types (e.g Convert timestamp from chr to datetime format)

Let's explore each edges and nodes!
:::

## [4.3]{style="color:#d496d4; background:none;"} Extracting the edges and nodes tables

First we need to extract the nodes and links tibble data frames using `as_tibble()` of **tibble** package package into two separate tibble dataframes called *nodes_tbl* and *edges_tbl* respectively.

```{r}
nodes_tbl <- as_tibble(graph_data$nodes)
edges_tbl <- as_tibble(graph_data$edges)
```

### [4.3.1]{style="color:#d496d4; background:none;"} Initial EDA {#EDA}

#### [4.3.1.1]{style="color:#d496d4; background:none;"} Check how many distinct nodes

```{r}
#| code-fold: true
table(nodes_tbl$type)
```

```{r}
#| code-fold: true
kable(head(nodes_tbl))
```

#### [4.3.1.2]{style="color:#d496d4; background:none;"} Check how many distinct edges

```{r}
#| code-fold: true
table(edges_tbl$type)
```

```{r}
#| code-fold: true
kable(head(edges_tbl))
```

#### [4.3.1.3]{style="color:#d496d4; background:none;"} Explore frequency distribution of nodes_tbl

In the code chunk below, `ExpCatViz()` of SmartEDA package is used to reveal the frequency distribution of all categorical fields in *nodes_tbl* dataframe.

```{r}
#| code-fold: true
#| fig-height: 26
#| fig-width: 10
ExpCatViz(nodes_tbl, col = "lightblue",margin=2,Page = c(7,2))
```

::: {.nursebox .nurse data-latex="nurse"}
#### **Observation** {#obsv1 .no-top-padding}

-   Node type is dominated Events (69%) followed by Relationship (25%) and Entity (6%).
-   We need to convert some of the data types for date and time from chr to datetime format.
-   High Missing Values Across Most Fields. Fields like reference, authority_level, and report_type are nearly all NA.
-   Fields like monitoring_type, assessment_type, movement_type, etc. also show very sparse use. Which supports the earlier [glimpse()](#glimpse) finding where most attribute fields are only relevant to certain subtypes (e.g. type = "events" and sub_type = "Communications" will have timestamp and content but other type and sub_types doesn't have these information). We need to conditionally filter or visualize based on both type and sub_type to avoid misleading analysis. For example, don’t rely on type: assessment_type unless sub_type == "Assessment".
:::

#### [4.3.1.4]{style="color:#d496d4; background:none;"} Explore frequency distribution of edges_tbl

Now we will reveal the frequency distribution of all categorical fields in *edges_tbl* dataframe.

```{r}
#| code-fold: true
ExpCatViz(data=edges_tbl,
          col="lightblue")
```

::: {.nursebox .nurse data-latex="nurse"}
#### **Observation** {.no-top-padding}

-   A large proportion (32%) of edges have missing/undefined type values (NA). But as we've discovered from the earlier findings in earlier [SmartEDA of nodes_tbl](#obsv1) and [glimpse()](#glimpse), we know that only certain type and sub_type have all information.

-   Significant portion of edges are evidence_for (32%), suggests more evidential or contextual links. Possibly connecting events or documents to people, vessels, or actions.

-   Communication events (e.g., who sent/received messages) only make up 18% each of the dataset means that communication-specific analyses should filter only these types, excluding NA and evidence_for.

-   These observations support the earlier suggestion in glimpse() to filter only communication related data for our visualisation later on.
:::

## [4.4]{style="color:#d496d4; background:none;"} Data Preparation {#dataprep}

After the initial [EDA](#EDA) done in the previous chapter, we will performs the following data cleaning and data wrangling tasks:

-   Extract type = "Event", and sub_type = "communication" events since we will be focusing on intercepted communications.
-   Convert timestamp from chr to datetime format.
-   Extract sender and receiver links.
-   Separate senders and receivers.
-   Join sender-receiver pairs.
-   Join with communication details.
-   Join to get sender/receiver labels.
-   Join to get entity type from sub_type.
-   Save the result for further analysis and visualisation usage.

### [4.4.1]{style="color:#d496d4; background:none;"} Extract communication events with timestamps

```{r}
comm_events <- nodes_tbl %>%
  filter(type == "Event", sub_type == "Communication")

kable(head(comm_events))
```

### [4.4.2]{style="color:#d496d4; background:none;"} Extract necessary fields

We'll use `id`, `timestamp` and `content`

```{r}
comm_events <- comm_events %>%
  select(id, timestamp = timestamp, content = content)

kable(head(comm_events))
```

### [4.4.3]{style="color:#d496d4; background:none;"} Convert timestamp from chr to datetime format

```{r}
comm_events <- comm_events %>%
  mutate(timestamp = ymd_hms(timestamp))

kable(head(comm_events))
```

### [4.4.3]{style="color:#d496d4; background:none;"} Filter edges for sender and receiver

```{r}
comm_edges <- edges_tbl %>%
  filter(type %in% c("sent", "received")) %>%
  select(source, target, type)

kable(head(comm_edges))
```

### [4.4.4]{style="color:#d496d4; background:none;"} Separate and rename

```{r}
senders <- comm_edges %>%
  filter(type == "sent") %>%
  rename(sender = source, message_id = target)

receivers <- comm_edges %>%
  filter(type == "received") %>%
  rename(receiver = target, message_id = source)
```

### [4.4.5]{style="color:#d496d4; background:none;"} Join sender and receiver info

```{r}
comm_pairs <- inner_join(senders, receivers, by = "message_id")

kable(head(comm_pairs))
```

### [4.4.6]{style="color:#d496d4; background:none;"} Combine with communication details

```{r}
comm_full <- comm_pairs %>%
  left_join(comm_events, by = c("message_id" = "id"))

kable(head(comm_full))
```

### [4.4.7]{style="color:#d496d4; background:none;"} Join entity labels for sender and receiver

```{r}
entity_labels <- nodes_tbl %>%
  filter(type == "Entity") %>%
  select(id, label, sub_type)

kable(head(entity_labels))
```

### [4.4.7]{style="color:#d496d4; background:none;"} Combine everything

```{r}
comm_full <- comm_full %>%
  left_join(entity_labels, by = c("sender" = "id")) %>%
  rename(sender_label = label, sender_type = sub_type) %>%
  left_join(entity_labels, by = c("receiver" = "id")) %>%
  rename(receiver_label = label, receiver_type = sub_type)

kable(head(comm_full))
```

### [4.4.8]{style="color:#d496d4; background:none;"} Derive weekday and hour of the day fields

Step 1: Deriving *weekday* and *hour of day* fields

Create two new fields namely *weekday* and *hour*. In this step, we will write a function to perform the task.

```{r}
#| code-fold: true
make_hr_wkday <- function(timestamp) {
  real_times <- ymd_hms(timestamp)
  dt <- data.table(
    weekday = weekdays(real_times),
    hour = hour(real_times)
  )
  return(dt)
}
```

Step 2: Deriving the communication tibble data frame

```{r}
weekday_levels <- c("Sunday", "Saturday", "Friday", 
                    "Thursday","Wednesday","Tuesday",  "Monday")

comm_full <- comm_full %>%
  mutate(
    date = as.Date(timestamp),
    weekday = factor(weekdays(timestamp), levels = weekday_levels),
    hour = factor(hour(timestamp), levels = 0:23),
    week = isoweek(timestamp),
    week_label = case_when(
      isoweek(timestamp) == min(isoweek(timestamp)) ~ "Week 1",
      isoweek(timestamp) == max(isoweek(timestamp)) ~ "Week 2",
      TRUE ~ "Other")
  )
```

::: callout-note
Beside extracting the necessary data into *attacks* data frame, `mutate()` of **dplyr** package is used to convert *weekday* and *hour* fields into **factor** so they'll be ordered when plotting
:::

Table below shows the tidy tibble table after processing.

```{r}
kable(head(comm_full))
```

### [4.4.9]{style="color:#d496d4; background:none;"} Save to CSV

For the purpose of this Take Home Exercise, we will extract out to CSV so the rest of the group can work with this cleaned data!

```{r}
#| eval: false
write_csv(comm_full, "data/communications_full.csv")
```

# [5]{style="color:#d496d4; background:none;"} Data Visualisation, Observation, and Insights

For **part 1a and 1b of the Mini Challenge 3** Let's look specifically into event nodes where communications occur.

## [5.1]{style="color:#d496d4; background:none;"} Develop a graph-based visual analytics approach to identify any daily temporal patterns in communications.

```{r}
comm_full %>%
  count(date) %>%
  ggplot(aes(x = date, y = n)) +
  geom_line(color = "steelblue", size = 1.2) +
  labs(
    title = "Daily Communication Volume Over 2 Weeks",
    x = "Date", y = "Number of Messages"
  ) +
  theme_minimal()
```

```{r}
comm_full %>%
  count(week_label, date) %>%
  ggplot(aes(x = date, y = n)) +
  geom_line(color = "steelblue", size = 1.2) +
  facet_wrap(~ week_label, scales = "free_x", ncol = 1, labeller = label_value) +
  labs(
    title = "Daily Communication Volume Per Week",
    x = "Date", y = "Number of Messages"
  ) +
  theme_minimal()
```

```{r}
comm_hourly_weekly <- comm_full %>%
  count(week_label, weekday, hour)

ggplot(comm_hourly_weekly, aes(x = hour, y = weekday, fill = n)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(option = "plasma", name = "Message Count") +
  labs(title = "Communication Intensity by Hour and Weekday",
       x = "Hour of Day", y = "Day of Week") +
  facet_wrap(~ week_label, ncol = 1) +
  theme_minimal()

```

::: {.nursebox .nurse data-latex="nurse"}
#### 🎀 **How do these patterns shift over the two weeks of observations?** {.no-top-padding}

-   In Week 1, daily messages volume remains relatively stable. This suggests routine or planned activity, likely related to operational coordination or normal business across the island.

-   However, in Week 2, the communication pattern becomes more volatile:

    -   There’s a sharp drop at the start of the week (Oct 08–09), possibly indicating reduced activity or deliberate radio silence.

    -   This is followed by a sudden spike on Oct 11 which is day 1 of the closure of Nemo Reef lasting 3 days up to Oct 13.

    -   The remainder of the week shows a steady decline.

        -   Furhter investigation done during this period and found messages on 14 Oct from Davis to Boss with content "Operation complete at Nemo Reef" and Davis to Mako with content "Well done on Nemo Reef operation completion" which may reflect completed operations or deliberate scaling back of activity.

-   The hourly heatmap supports this shift:

    -   In Week 1, communication is consistently concentrated during specific hours of the day (e.g., mornings or early evenings), reflecting regular daily routines.

    -   In Week 2, these temporal patterns shift, with more dispersed or intensified communication on certain days — especially around the spike, indicating possible emergency coordination or reactive behavior.
:::

## [5.2]{style="color:#d496d4; background:none;"} Focus on a specific entity and use this information to determine who has influence over them.

For **part 1c of the Mini Challenge 3** we will be looking at the influence over the nodes.

**Influence** could be indicated by:

-   Frequent communication to or from the target

-   One-way communication (many messages received, few sent)

-   Connected intermediaries (network centrality or shared contacts)

Let's look into the network graph:

### [5.2.1]{style="color:#d496d4; background:none;"} Communication network graph

```{r}
# Create edge list
graph_edges <- comm_full %>%
  count(sender_label, receiver_label) %>%
  filter(!is.na(sender_label), !is.na(receiver_label)) %>%
  rename(from = sender_label, to = receiver_label, value = n)

# Create node list from edges
graph_nodes <- tibble(name = unique(c(graph_edges$from, graph_edges$to))) %>%
  mutate(id = name, label = name, group = name)

# visNetwork expects numeric node IDs, but we can use labels for both
visNetwork(nodes = graph_nodes, edges = graph_edges) %>%
  visEdges(arrows = "to") %>%
  visOptions(highlightNearest = TRUE, nodesIdSelection = TRUE) %>%
  visLayout(randomSeed = 123) %>%
  visPhysics(stabilization = TRUE) %>%
  visInteraction(navigationButtons = TRUE) %>%
  visLegend()

```

Deep dive

#### [5.2.1.1]{style="color:#d496d4; background:none;"} receiver_type: person and sender_type: person

```{r}
# Create edge list
graph_edges <- comm_full %>%
  filter(receiver_type == "Person" & sender_type == "Person") %>%
  count(sender_label, receiver_label) %>%
  filter(!is.na(sender_label), !is.na(receiver_label)) %>%
  rename(from = sender_label, to = receiver_label, value = n)

# Create node list from edges
graph_nodes <- tibble(name = unique(c(graph_edges$from, graph_edges$to))) %>%
  mutate(id = name, label = name, group = name)

# visNetwork expects numeric node IDs, but we can use labels for both
visNetwork(nodes = graph_nodes, edges = graph_edges) %>%
  visEdges(arrows = "to") %>%
  visOptions(highlightNearest = TRUE, nodesIdSelection = TRUE) %>%
  visLayout(randomSeed = 123) %>%
  visPhysics(stabilization = TRUE) %>%
  visInteraction(navigationButtons = TRUE) %>%
  visLegend()

```

#### [5.2.1.2]{style="color:#d496d4; background:none;"} receiver_type: person and sender_type: vessel

```{r}
# Create edge list
graph_edges2 <- comm_full %>%
  filter(receiver_type == "Person" & sender_type == "Vessel") %>%
  count(sender_label, receiver_label) %>%
  filter(!is.na(sender_label), !is.na(receiver_label)) %>%
  rename(from = sender_label, to = receiver_label, value = n)

# Create node list from edges
graph_nodes <- tibble(name = unique(c(graph_edges2$from, graph_edges2$to))) %>%
  mutate(id = name, label = name, group = name)

# visNetwork expects numeric node IDs, but we can use labels for both
visNetwork(nodes = graph_nodes, edges = graph_edges2) %>%
  visEdges(arrows = "to") %>%
  visOptions(highlightNearest = TRUE, nodesIdSelection = TRUE) %>%
  visLayout(randomSeed = 123) %>%
  visPhysics(stabilization = TRUE) %>%
  visInteraction(navigationButtons = TRUE) %>%
  visLegend()

```

#### [5.2.1.3]{style="color:#d496d4; background:none;"} receiver_type: vessel and sender_type: person

```{r}
# Create edge list
graph_edges2 <- comm_full %>%
  filter(receiver_type == "Vessel" & sender_type == "Person") %>%
  count(sender_label, receiver_label) %>%
  filter(!is.na(sender_label), !is.na(receiver_label)) %>%
  rename(from = sender_label, to = receiver_label, value = n)

# Create node list from edges
graph_nodes <- tibble(name = unique(c(graph_edges2$from, graph_edges2$to))) %>%
  mutate(id = name, label = name, group = name)

# visNetwork expects numeric node IDs, but we can use labels for both
visNetwork(nodes = graph_nodes, edges = graph_edges2) %>%
  visEdges(arrows = "to") %>%
  visOptions(highlightNearest = TRUE, nodesIdSelection = TRUE) %>%
  visLayout(randomSeed = 123) %>%
  visPhysics(stabilization = TRUE) %>%
  visInteraction(navigationButtons = TRUE) %>%
  visLegend()

```

#### [5.2.1.4]{style="color:#d496d4; background:none;"} receiver_type: organisation and sender_type: person

```{r}
# Create edge list
graph_edges2 <- comm_full %>%
  filter(receiver_type == "Organization" & sender_type == "Person") %>%
  count(sender_label, receiver_label) %>%
  filter(!is.na(sender_label), !is.na(receiver_label)) %>%
  rename(from = sender_label, to = receiver_label, value = n)

# Create node list from edges
graph_nodes <- tibble(name = unique(c(graph_edges2$from, graph_edges2$to))) %>%
  mutate(id = name, label = name, group = name)

# visNetwork expects numeric node IDs, but we can use labels for both
visNetwork(nodes = graph_nodes, edges = graph_edges2) %>%
  visEdges(arrows = "to") %>%
  visOptions(highlightNearest = TRUE, nodesIdSelection = TRUE) %>%
  visLayout(randomSeed = 123) %>%
  visPhysics(stabilization = TRUE) %>%
  visInteraction(navigationButtons = TRUE) %>%
  visLegend()

```

::: {.nursebox .nurse data-latex="nurse"}
#### 🎀 **Who has influence over them?** {.no-top-padding}

From the network graph, we observe these top affiliated entities based on communication frequency, categorized by type. Further frequency check was also conducted using pivot table in excel using the CSV exported earlier:

-   Organization:
    -   Green Guardian (Organization) has the highest number of communication sent, they are the local conservationist group who is central to the Nemo Reef conflict.

    -   Top 2 receiver and sender is actually Oceanus City Council which is a key governing body, possibly linked to expedited approvals.

    -   V. Miesel Shipping seems to be the potential marine logistic partner for this illegal activity.
-   Vessels
    -   Mako (Vessel) has the highest number of communication received , likely crucial in coordination or transport.

    -   Reef Guardian — potentially linked with conservation, possibly under Green Guardians.

    -   Remora (62 messages), Neptune (60), Sentinel (41), Horizon (36), EcoVigil (29) — may play roles in patrol, logistics, or covert operations.
-   Persons
    -   “Boss” appears to be high-ranking, managing flow between entities like Mrs. Money (possibly financial operator) and Middleman (coordinator). Their involvement with Mako and shipping entities implies they may be central to illicit logistics or moving resources/people covertly.

    -   “The Intern” is likely a junior operative or data/communications relay, possibly linked directly to Mrs. Money. Their mirrored exchanges with “The Lookout” point toward coordinated surveillance or monitoring tasks.

    -   “The Lookout” acts as a watcher/spotter, deeply involved in tracking activity around conservation zones. Their tie to Sam and Green Guardians might indicate a dual role or impersonation within the environmental group.

    -   We observe that Davis is receiving messages from or sending to "Boss", he also observed giving instructions to Mako (a high-traffic vessel). It's plausible that Davis may play a hands-on operational role, where he is looped into the command chain, receives instructions from Boss, relays or enforces those orders to operatives (e.g., Mako, vessel contacts). Further investigation done in the excel CSV exported earlier and found that Event_Communication_343 from Davis revealed in the content that he is Captain Davis who is in charge of Mako.

    -   Liam Thorne seems to be a double agent who is officially working for Oceanus City Council but reporting to the illegal group as an informant (this is confirmed after reading through the detail communication contents), facilitating or delay approvals, making him crucial in the closure of Nemo Reef or the Sailor Shift project.
-   Locations
    -   Himark Harbor (45 messages), Paackland Harbor (32 messages) — logistical hubs or strategic points.
:::

## [5.3]{style="color:#d496d4; background:none;"} Pseudonym Usage and The Reflection on Activity

For **part 3c of the Mini Challenge 3** we will be reflecting on what have been uncovered earlier.

::: {.nursebox .nurse data-latex="nurse"}
#### 🎀 **How does our understanding of activities change given our understanding of pseudonyms?** {.no-top-padding}

-   Before Identifying Pseudonyms
    -   The dataset looked like a mix of bureaucratic, logistical, and informal communications — possibly between officials, vessels, and environmental actors. The presence of names like “Boss” or “The Lookout” hinted at secrecy, but without clear roles, their significance was ambiguous.
-   After Pseudonym Analysis
    -   Now that we’ve mapped likely roles behind pseudonyms, the story reveals covert orchestration, manipulation of governance, and strategic masking of identities:
        -   A Coordinated Inner Circle Emerges Boss appears to be a top-level orchestrator, managing directives via trusted intermediaries (e.g., The Middleman, Mrs. Money).
        -   The Intern and The Lookout serve specific operational functions as mentioned in the previous obserfaction
        -   Real Individuals Are Embedded in the Scheme. Captain Davis is coordinating the operational for Vessel Mako and receiving instructions from Boss. His seemingly bureaucratic or neutral position is now suspect as central to execution.
        -   Liam Thorne appears to be straddling official (City Council) and unofficial channels, indicating a mole or power-broker ensuring approvals pass smoothly. This suggests institutional corruption, not just individual misbehavior.

We now know that the Nemo Reef closure not as a public safety or ecological move but as a coordinated action likely benefiting Sailor Shift's team (filming) or facilitating illegal transitions (e.g., laundering operations into tourism). The expedited logistics and vessel movements are also calculated move that may represent covert transport or cover-up operations.
:::

# [6]{style="color:#d496d4; background:none;"} Reference

\- [**VAST 2024 Mini Challenge 3**](https://vast-challenge.github.io/2025/MC3.html){target="_blank"}\
- [**jsonlite**](https://cran.r-project.org/package=jsonlite){target="_blank"}: A robust and high-performance JSON parser and generator for R, useful for working with web APIs and structured data. \
- [**tibble**](https://cran.r-project.org/package=tibble){target="_blank"}: A modern reimagining of data frames in R that prints cleaner and is better for programming.\
- [**dplyr**](https://cran.r-project.org/package=dplyr){target="_blank"}: A grammar of data manipulation, making data wrangling concise, fast, and readable.\
- [**knitr**](https://cran.r-project.org/package=knitr){target="_blank"}: A dynamic report generation tool in R, used for rendering markdown reports and tables.\
- [**SmartEDA**](https://cran.r-project.org/package=SmartEDA){target="_blank"}: Automatically generates EDA reports including statistical summaries and visualizations.\
- [**lubridate**](https://cran.r-project.org/package=lubridate){target="_blank"}: Simplifies the parsing, manipulation, and extraction of date-time data.\
- [**readr**](https://cran.r-project.org/package=readr){target="_blank"}: Provides fast and friendly functions for reading and writing rectangular data (e.g., CSV).\
- [**ggplot2**](https://cran.r-project.org/package=ggplot2){target="_blank"}: A versatile and widely used R package for creating elegant data visualizations based on the Grammar of Graphics.\
- [**visNetwork**](https://datastorm-open.github.io/visNetwork/){target="_blank"}: Enables interactive network visualizations in R using the vis.js JavaScript library.\
- [**tidygraph**](https://cran.r-project.org/package=tidygraph){target="_blank"}: Provides a tidy framework for network analysis, integrating graph theory with tidy data principles.\
- [**ggraph**](https://cran.r-project.org/package=ggraph){target="_blank"}: Extends ggplot2 to visualize network and graph structures with customizable layouts.\
- [**ggrepel**](https://cran.r-project.org/package=ggrepel){target="_blank"}: Automatically adjusts overlapping text labels in ggplot2 and ggraph visualizations to improve clarity.

# [7]{style="color:#d496d4; background:none;"} Appendix

## [7.1]{style="color:#d496d4; background:none;"} Appendix A {#appendixa}

![](data/MC3_data_description.pdf){fig-align="center" width="700" height="350"}
