---
title: "Creating Ternary Plot with R"
author: "Vanessa Riadi"
date: "May 8, 2025"
date-modified: "last-modified"
format:
  html:
    toc: true
execute: 
  echo: true
  eval: true
  warning: false
  freeze: true
---

# [1]{style="color:#d496d4; background:none;"} Overview

[Ternary plots](https://en.wikipedia.org/wiki/Ternary_plot) are a way of displaying the distribution and variability of three-part compositional data. (For example, the proportion of aged, economy active and young population or sand, silt, and clay in soil.) It's display is a triangle with sides scaled from 0 to 1. Each side represents one of the three components. A point is plotted so that a line drawn perpendicular from the point to each leg of the triangle intersect at the component values of the point.

In this hands-on, you will learn how to build ternary plot programmatically using R for visualising and analysing population structure of Singapore.

The hands-on exercise consists of four steps:

-   Install and launch **tidyverse** and **ggtern** packages.
-   Derive three new measures using *mutate()* function of **dplyr** package.
-   Build a static ternary plot using *ggtern()* function of **ggtern** package.
-   Build an interactive ternary plot using *plot-ly()* function of **Plotly R** package.

# [2]{style="color:#d496d4; background:none;"} Installing and launching R packages

Two main R packages will be used:

-   [**ggtern**](http://www.ggtern.com/), a ggplot extension specially designed to plot ternary diagrams. The package will be used to plot static ternary plots.
-   [**Plotly R**](https://plot.ly/r/), an R package for creating interactive web-based graphs via plotly's JavaScript graphing library, plotly.js . The **plotly R** libary contains the *ggplotly* function, which will convert **ggplot2** figures into a Plotly object.

We will also need to ensure that selected **tidyverse** family packages namely: **readr**, **dplyr** and **tidyr** are also installed and loaded.

::: {.nursebox .nurse data-latex="nurse"}
The new version of [**ggtern 	3.5.0**](https://cran.r-project.org/web/packages/ggtern/index.html) package is compatible to the **ggplot2 version 3.5.0 and above**.
:::

The code chunks below will accomplish the task.

```{r}
pacman::p_load(plotly, ggtern, tidyverse)
```

# [3]{style="color:#d496d4; background:none;"} Data Preparation

## [3.1]{style="color:#d496d4; background:none;"} The data

We will use the [Singapore Residents by Planning AreaSubzone, Age Group, Sex and Type of Dwelling, June 2000-2018](https://www.singstat.gov.sg/find-data/search-by-theme/population/geographic-distribution/latest-data) data will be used. The downloaded file is a csv format with filename *respopagsex2000to2018_tidy.csv*.

## [3.2]{style="color:#d496d4; background:none;"} Importing Data

Import data using ***read_csv()*** function of **readr** package.

```{r}
#| echo: true
#| code-fold: true
pop_data <- read_csv("../data/respopagsex2000to2018_tidy.csv") 
```

## [3.3]{style="color:#d496d4; background:none;"} Preparing the Data

Next, use the ***mutate()*** function of **dplyr** package to derive three new measures, namely: young, active, and old.

```{r}
#| echo: true
agpop_mutated <- pop_data %>%
  mutate(`Year` = as.character(Year))%>%
  spread(AG, Population) %>%
  mutate(YOUNG = rowSums(.[4:8]))%>%
  mutate(ACTIVE = rowSums(.[9:16]))  %>%
  mutate(OLD = rowSums(.[17:21])) %>%
  mutate(TOTAL = rowSums(.[22:24])) %>%
  filter(Year == 2018)%>%
  filter(TOTAL > 0)
```

# [4]{style="color:#d496d4; background:none;"} Plotting Ternary Diagram with R

[Ternary Plot Lesson Slide](https://isss608-ay2024-25apr.netlify.app/lesson/lesson05/lesson05-vma#/ternary-plot)

::: {.nursebox .nurse data-latex="nurse"}
ggtern themes have been designed to mimic the default themes available within ggplot2. 

Themes available such as ggtern, rgbg, rgbw, darker, gray, bw, classic, minimal, linedraw, light, void and dark. 
:::

## [4.1]{style="color:#d496d4; background:none;"} Plotting a static ternary diagram

Use ***ggtern()*** function of **ggtern** package to create a simple ternary plot.

```{r}

#Building the static ternary plot
ggtern(data=agpop_mutated,aes(x=YOUNG,y=ACTIVE, z=OLD)) +
  geom_point()
```

```{r}

#Building the static ternary plot
ggtern(data=agpop_mutated, aes(x=YOUNG,y=ACTIVE, z=OLD)) +
  geom_point() +
  labs(title="Population structure, 2015") +
  theme_rgbw()
```

## [4.2]{style="color:#d496d4; background:none;"} Plotting an interative ternary diagram

The code below create an interactive ternary plot using ***plot_ly()*** function of **Plotly R**.

::: panel-tabset
## 🎀 [PRACTICE]{style="color:#d496d4"}
```{r}
#| echo: false

# ---------- helper functions ----------

label <- function(txt) {
  list(
    text = txt, x = 0.1, y = 1, ax = 0, ay = 0,
    xref = "paper", yref = "paper", align = "center",
    font       = list(family = "serif", size = 15, color = "white"),
    bgcolor    = "#b3b3b3",
    bordercolor = "black",
    borderwidth = 2
  )
}

axis <- function(txt, col) {
  list(
    title    = list(text = txt, font = list(color = col)),
    tickformat = ".0%",
    tickfont = list(size = 10, color = col),
    linecolor = col                 # axis line
  )
}

ternaryAxes <- list(
  aaxis = axis("Young" , "#0033A0"),   # red
  baxis = axis("Active", "#86110D"),   # green
  caxis = axis("Old"   , "#1C6B17")    # blue
)

# ---------- interactive plot ----------

plot_ly(
  agpop_mutated,
  a    = ~YOUNG,
  b    = ~ACTIVE,
  c    = ~OLD,
  type = "scatterternary",
  mode = "markers",
  color = I("black"),

  # nice, readable tooltip
  hovertemplate = paste(
    "Young:  %{a:.0%}",
    "<br>Active: %{b:.0%}",
    "<br>Old:    %{c:.0%}",
    "<extra></extra>"
  )
) %>%
  layout(
    title       = list(text = "Population structure, 2015"),
    annotations = label("Ternary Markers"),
    ternary     = ternaryAxes,
    margin      = list(t = 90)
  )


```
```{r}
#| eval: false

# ---------- helper functions ----------

label <- function(txt) {
  list(
    text = txt, x = 0.1, y = 1, ax = 0, ay = 0,
    xref = "paper", yref = "paper", align = "center",
    font       = list(family = "serif", size = 15, color = "white"),
    bgcolor    = "#b3b3b3",
    bordercolor = "black",
    borderwidth = 2
  )
}

axis <- function(txt, col) {
  list(
    title    = list(text = txt, font = list(color = col)),
    tickformat = ".0%",
    tickfont = list(size = 10, color = col),
    linecolor = col                 # axis line
  )
}

ternaryAxes <- list(
  aaxis = axis("Young" , "#0033A0"),   # red
  baxis = axis("Active", "#86110D"),   # green
  caxis = axis("Old"   , "#1C6B17")    # blue
)

# ---------- interactive plot ----------

plot_ly(
  agpop_mutated,
  a    = ~YOUNG,
  b    = ~ACTIVE,
  c    = ~OLD,
  type = "scatterternary",
  mode = "markers",
  color = I("black"),

  # nice, readable tooltip
  hovertemplate = paste(
    "Young:  %{a:.0%}",
    "<br>Active: %{b:.0%}",
    "<br>Old:    %{c:.0%}",
    "<extra></extra>"
  )
) %>%
  layout(
    title       = list(text = "Population structure, 2015"),
    annotations = label("Ternary Markers"),
    ternary     = ternaryAxes,
    margin      = list(t = 90)
  )


```
## Given Code
```{r}
#| echo: false

# reusable function for creating annotation object
label <- function(txt) {
  list(
    text = txt, 
    x = 0.1, y = 1,
    ax = 0, ay = 0,
    xref = "paper", yref = "paper", 
    align = "center",
    font = list(family = "serif", size = 15, color = "white"),
    bgcolor = "#b3b3b3", bordercolor = "black", borderwidth = 2
  )
}

# reusable function for axis formatting
axis <- function(txt) {
  list(
    title = txt, tickformat = ".0%", tickfont = list(size = 10)
  )
}

ternaryAxes <- list(
  aaxis = axis("Young"), 
  baxis = axis("Active"), 
  caxis = axis("Old")
)

# Initiating a plotly visualization 
plot_ly(
  agpop_mutated, 
  a = ~YOUNG, 
  b = ~ACTIVE, 
  c = ~OLD, 
  color = I("black"), 
  type = "scatterternary",
  mode = "markers"
) %>%
  layout(
    annotations = label("Ternary Markers"), 
    ternary = ternaryAxes
  ) 
```

```{r}
#| eval: false

# reusable function for creating annotation object
label <- function(txt) {
  list(
    text = txt, 
    x = 0.1, y = 1,
    ax = 0, ay = 0,
    xref = "paper", yref = "paper", 
    align = "center",
    font = list(family = "serif", size = 15, color = "white"),
    bgcolor = "#b3b3b3", bordercolor = "black", borderwidth = 2
  )
}

# reusable function for axis formatting
axis <- function(txt) {
  list(
    title = txt, tickformat = ".0%", tickfont = list(size = 10)
  )
}

ternaryAxes <- list(
  aaxis = axis("Young"), 
  baxis = axis("Active"), 
  caxis = axis("Old")
)

# Initiating a plotly visualization 
plot_ly(
  agpop_mutated, 
  a = ~YOUNG, 
  b = ~ACTIVE, 
  c = ~OLD, 
  color = I("black"), 
  type = "scatterternary",
  mode = "markers"
) %>%
  layout(
    annotations = label("Ternary Markers"), 
    ternary = ternaryAxes
  ) 
```
:::

# [5]{style="color:#d496d4; background:none;"} References

## [5.1]{style="color:#d496d4; background:none;"} Readings

### [5.1.1]{style="color:#d496d4; background:none;"} Core Readings

-   [An Introduction to Visual Multivariate Analysis](http://www.perceptualedge.com/articles/b-eye/visual_multivariate_analysis.pdf)
-   [Ternary Plot](https://en.wikipedia.org/wiki/Ternary_plot)
-   [Multivariate Data Glyphs](https://link-springer-com.libproxy.smu.edu.sg/content/pdf/10.1007%2F978-3-540-33037-0_8.pdf)

### [5.1.2]{style="color:#d496d4; background:none;"} Optional Readings

-   [The scatter-plot matrix: a great tool](http://junkcharts.typepad.com/junk_charts/2010/06/the-scatterplot-matrix-a-great-tool.html)
-   [Ternary Diagram](https://www.sciencedirect.com/topics/engineering/ternary-diagram)
-   [Reading a Ternary Diagram](http://csmgeo.csm.jmu.edu/geollab/fichter/Sedrx/readternary.html)
-   [Multidimensional Detective](http://www.ifs.tuwien.ac.at/~silvia/wien/vu-infovis/articles/Inselberg-InfoVis97.pdf)

## [5.2]{style="color:#d496d4; background:none;"} R Packages

-   [**ggtern**](http://www.ggtern.com/), a ggplot extension specially designed to plot ternary diagrams. The package will be used to plot static ternary plots.
-   [**Plotly R**](https://plot.ly/r/), an R package for creating interactive web-based graphs via plotly's JavaScript graphing library, plotly.js . The **plotly R** libary contains the *ggplotly* function, which will convert **ggplot2** figures into a Plotly object.

